---
title: 'Week 3: Functions and Packages'
output: html_document
---

# Functions

Welcome to Week 3! In this tutorial, you'll learn all about functions in R. This includes building your own functions, using R's built-in functions, and loading packages that contain helpful tools.

If you have any questions about the contents of this tutorial, please post them in the `Week 3 Pre-Class Tutorial` thread on Ed.

## Function? I sure hope it does.

Functions allow us to take an input, run some code, and send back an output. In fact, if you've ever taken a math class before, you've already seen functions!

$$
f(x) = 2x
$$

Look familiar? This function is very simple. It takes an input $x$, multiplies it by two, and gives back the answer. You can run it over and over again for different values:

$$
f(1)=2*1=2\\
f(-10)=2*-10=-20\\
f(0)=2*0=0
$$

R lets us do the same exact thing. We can provide instructions inside a function that we use later for different inputs. R uses `function(){}` to define these inputs and instructions. The inputs go inside the `()` and the instructions go inside the `{}`. In this case, our input `x` goes inside the parentheses, and our instructions `2*x` goes inside the curly brackets. We give the function the name `times_two` so we can reference it later. So if we want to turn our math function into an R function called `times_two`, we can say:

```{r}
times_two = function(x){
  2*x
}
```

After we run this code and our function is saved, we can **call** to the function using its name followed by our input in parenthesis. For `times_two`, all we have to do is give `x` a value:

```{r}
times_two(x=1)
times_two(x=-10)
times_two(x=0)
```

Each time we run the function, it **returns** back the last thing it calculated. For this example, it only calculates `2*x`, so that is what it returns.

We can also use a `return()` statement to explicitly state what we want the function output to be. A `return()` statement tells R that we want to exit the function, and any code below it won't be run. 

```{r}
times_two = function(x){
  return(2*x)
  x = x + 1 # This code will never be reached
  x
}

times_two(x=4)
```

### Practice Problem

Write your own function called `times_four_plus_one` that takes one input `x`, multiplies it by four, and adds one. Then, call your function three times with different inputs to make sure it works right.

```{r}
...
```

## Getting more complex

Functions can contain any instructions, not just math. Let's write some code that will calculate the absolute value of a number:

```{r}
value = -2

if (value < 0) {
  value * -1
} else {
  value
}
```

Writing all those lines out every time we want an absolute value is a bit of a pain. Instead, we can turn that into a function:

```{r}
absolute_value = function(x) {
  if (x < 0) {
    x * -1
  } else {
    x
  }
}

absolute_value(x=-2)
absolute_value(x=10)
```

Let's break down what happened here:

1.  We created a function called `absolute_value` that has one input `x` inside the parentheses. `x` has not been assigned a value, so we will have to do that when we call the function. We then put our `if` statement, indented to the right, inside the curly brackets.
2.  We called the function using its name followed by parenthesis. `x` needed a value, so we gave it a value inside the parenthesis.
3.  For `x=-2`, `x` is less than 0, so the `if` statement is `TRUE`, running the first line `x * -1`. Because the `if` statement was `TRUE`, the `else` line is not run. Because `x * -1` was the last thing the function calculated, it returns `x * -1` or 2.
4.  For `x=10`, `x` is not less than 0, so the `if` statement is `FALSE`, which skips straight to `else`. The `else` line is simply `x`. Because `x` was the last thing the function calculated, it returns `x` or 10.

In this way, we can walk through, step by step, what every function does. However, we do not have to be limited by only one input. Functions can accept many inputs, each with their own name. Here's an addition equation generator that accepts two inputs and prints the addition equation for the two:

```{r}
# Accepts two numbers
# Returns a character with the formula for the sum
generate_addition = function(num1, num2){
  paste0(num1, "+", num2, "=", num1+num2)
} 

generate_addition(num1=45, num2=22)
```

It's always a good idea to write out `num1=` and `num2=` just to make sure you're feeding the right values into the right inputs. However, you can also leave off the `num1=` and `num2=` as long as you put the inputs in the right order.

```{r}
generate_addition(45, 22)
generate_addition(22, 45)
```

Also note that no `num1` or `num2` variables appeared in your "Environment" tab in the top right of your screen. Variables created inside a function stay inside the function. If you want to access something after your function runs, your function needs to return it by having it be the last thing the function calculates.

```{r}
# Calculates the variable 'sum', but 'sum' is not run and gets thrown away after the function runs
func_that_does_nothing = function(num1, num2){
  sum = num1 + num2
}
# Calculates 15 and immediately throws it out
func_that_does_nothing(num1=5, num2=10)

# Calculates the variable 'sum' and runs 'sum' before it gets thrown away after the function runs
func_that_works = function(num1, num2){
  sum = num1 + num2
  sum
}
# Calculates 90 and returns it before throwing it out
func_that_works(num1=15, num2=75)
```

### Practice Problem

Write your own `generate_multiplication` function that accepts two numbers and prints the multiplication equation for the two. Call your function a few times to make sure it's working right.

```{r}
...
```

Wow! That's a lot of work. If only someone had already written these functions for us...

## Built-in Functions

R has a bunch of built-in functions that help us get our work done. In fact, you have been using functions this whole time! `data.frame()` is a useful function that we learned about last week. And during week one, we learned about a bunch of math functions like `exp()`, `log()`, and even `abs()` which gives us the absolute value!

An especially useful built-in function is `seq()`. This function allows you to generate a sequence of numbers. Now, we already learned that the colon `:` can be used to generate numbers in order. `seq()` expands on this by taking three inputs: `from=`, `to=`, and `by=`. `from` tells it our starting number, `to` tells it our maximum number, and `by` tells it the gaps between the numbers. If we want the numbers 1 through 10, we can just say `1:10` or `seq(from=1, to=10, by=1)`. However, if we want every odd number from 1 to 10, we cannot use the `:` and have to use `seq(from=1, to=10, by=2)`.

```{r}
# 1 through 10
1:10
seq(1, 10)
# odd numbers 1 through 10
seq(1, 10, 2)
```

### Practice Problem

Now you use `seq()` to generate all multiples of 0.2 from 0 to 5 (e.g. 0, 0.2, 0.4, ... , 4.8, 5).

```{r}
...
```

## Getting Help

R has a lot of functions built in, so it is almost impossible to remember all of them. Instead, we have the `?` operator. When you write this operator before a function name in R, it will pop up some useful information.

```{r}
?sample
```

When we run this line, a little window pops up on the right side describing what `sample()` does. We learn that it takes a sample of size `size` from the elements of `x`. So let's randomly pick 3 numbers between 1 and 10.

```{r}
sample(x=1:10, size=3)
```

There is also an optional input `replace` that is normally equal to `FALSE`. Further down under **Arguments**, we learn that `replace` tells it "should sampling be with replacement?". In other words, should the same element be able to be picked multiple times? If we want that to be yes, we can tell the function that. Try running this code multiple times until you get a repeated number.

```{r}
sample(x=1:10, size=3, replace=TRUE)
```

The help page on the right side also has examples down at the bottom to help us. You can also always use Google to help you figure out how to use these functions, but the `?` operator works even without Internet!

### Practice Problem

Use `?sample` (and the Internet if you need) to figure out how to sample 5 letters from "RRules" *with replacement*.

```{r}
...
```

**Note:** If your code just prints out "RRules" 5 times, you did not do it right.

## Any Number of Arguments

Sometimes, functions will have an input that just says `…`. For example:

```{r}
?max()
```

The ellipsis means that `max()` can accept any number of arguments and give you the maximum of them.

```{r}
max(1, 5, "B", 22)
```

### Practice Problem

Try it out! The `max()` and `min()` functions will give you the maximum and minimum, respectively, of numbers or letters. They will accept any number of inputs, including a pre-made vector. Try finding different minimums and maximums:

```{r}
min()
max()

some_numbers = c(100, 50, 29, 72458, 1, 0)
max(some_numbers)
```

# Packages

Writing functions sure is fun, but I don't want to have to write a function for everything in R. Luckily, many amazing data scientists have come before us, written useful functions, and stored them in packages. Packages are bits of code you download from the internet to use in your own code. In R, there are two steps to using a package. 1) Install the package with `install.packages()` and 2) Tell R you want to use that package with `library()`. One of the most useful packages is called tidyverse. This package contains a multitude of packages that help us import, clean, and visualize data. Let's download tidyverse.

```{r}
# Because we're working in R markdown, we have to tell it where to search for the package. That's why we put repos="https://cloud.r-project.org". If you run install.packages() in the console or in a normal R file, you don't need the repos= part.
install.packages("tidyverse", repos="https://cloud.r-project.org")
```

Note that we put the package name in quotes like a string. Because the package is not downloaded yet, we have to put it in quotes. Now that it is downloaded, we can tell R we are ready to use tidyverse. Because it is downloaded, it no longer needs to be in quotes because it exists on our computer.

```{r}
library(tidyverse)
```

### Practice Problem

Another useful package is called haven. Haven allows you to import SPSS, STATA, and SAS data files. Go ahead and follow the two steps to download and install haven.

```{r}
...
```

# Importing Data

By installing tidyverse, one of the packages that gets installed is called readr. This package gives us useful functions to import data into R because who wants to type out 3,000 rows??? Let's learn about importing data. Before you get started, **make sure you're working in your R Studio project that you made during class on Week 1**. That means this file should be saved in your `pre-class` folder. If you're not already doing that, you can click File\>Save As... and save it in your R project folder. The top right of your R Studio should say something like "EPI 202 R Class".

Go to Canvas and download the wage_data.csv file under Week 3. After it has downloaded, move `wage_data.csv` into your `data` folder in this R project. When you go into R, you should be able to click on the "Files" tab, and the `data` folder and see `wage_data.csv` in that folder.

Now we're ready to import the data. We will use the `read_csv()` function from the `readr` package. You can check out what `read_csv()` does using the `?` function. If you get an error, you did not download and library tidyverse yet.

```{r}
...
```

`read_csv()` has a bunch of options that lets us read in different files, but the only required input is the `file` we want in r. We give the file we want to import using a character. Because we are super organized and we are using an R Studio Project, all we have to say is look `here::here` in the `data` folder for a file called `wage_data.csv`. We do this by saying `here::here("data/wage_data.csv")`. You can even use the Tab key to autocomplete the name for you!

```{r}
read_csv(file=here::here("data", "wage_data.csv"))
```

*Why do we have to say `here::here()`*? This is the way tell R Markdown to look at the R Project for help finding the file. If you're working in a normal R file (one that ends with `.R` instead of `.Rmd)`, you don't have to do this.

### Practice Problem

Notice that the data were printed without it appearing in the top right "Environment" tab. That's because we didn't assign it a variable! Try importing the wage data again, this time assigning it to a variable called `wages`.

```{r}
# Write your code to import the wage data


# Print out the wages variable to check and make sure it was saved
wages
```
